---
title: "Reproducible Research Peer Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Examining Steps Taken Over Time

This markdown document contains code and output for Johns Hopkins Reproducible Research Coursera course for the week 2 peer reviewed assignment.  


##Loading and preprocessing the data (Question 1)

```{r}
#setting the working directory
rm(list=ls())
getwd()

setwd("C:/Users/Daniel Slyngstad/Documents/R Working Directory/RMOOC2/MOOC5")

#loading requisite packages
library(tidyverse)
library(psych)
library(lubridate)

#reading in the data
df <- read.csv('activity.csv', header = TRUE) 

#double checking the data format
head(df)
dim(df)
```


##What is mean total number of steps taken per day? (Question 2)

```{r}
#creating a dataset aggregated by date
bydate_df_sum_noNA <- aggregate(df[, 1], list(df$date), sum, na.rm=TRUE)
names(bydate_df_sum_noNA) <- c('day', 'total_steps')
```

###Plotting mean total steps per day
```{r}
#plotting average total steps taken each day as a histogram
hist(bydate_df_sum_noNA$total_steps, xlab = 'Day Number', main = 'Total Steps per Day', col = 10, breaks = 25)
```

###Checking the mean and median of total steps per day
```{r}
#mean and median total steps 
mean(bydate_df_sum_noNA$total_steps, na.rm=TRUE)
median(bydate_df_sum_noNA$total_steps, na.rm=TRUE)
```


##What is the average daily activity pattern? (Question 3)
```{r}
#creating an aggregated dataframe by interval
byinterval_df_noNA <- aggregate(df[, 1], list(df$interval), mean, na.rm=TRUE)
names(byinterval_df_noNA) <- c("interval", 'steps')

#double checking the aggregated data
head(byinterval_df_noNA)
dim(byinterval_df_noNA)
```


###Plotting as a time series
```{r}
plot(byinterval_df_noNA$steps, type = 'l', main = 'Steps Taken by Five Minute Interval', 
     xlab = '5 Minute Interval Number', ylab = 'Average Steps')
```

###Finding the max interval
```{r}
byinterval_df_noNA$interval[byinterval_df_noNA$steps == max(byinterval_df_noNA$steps)]
```


##Imputing missing values (Question 4)
```{r}
#calculating the number of missing i.e. FALSEs
table(complete.cases(df))

#verifying that this approach returns a boolean
is.na(df$steps[1])

#imputing means from interval level data into raw data
mean_imp_df <- df 
for (i in 1:nrow(mean_imp_df)) {
  if (is.na(mean_imp_df$steps[i])) {
    mean_imp_df$steps[i] <- byinterval_df_noNA[which(mean_imp_df$interval[i] == byinterval_df_noNA$interval), ]$steps
  }
}
```

###Plotting the results
```{r}
#plotting total steps per day with imputed missing values
bydate_df_sum_imputed <- aggregate(mean_imp_df[, 1], list(df$date), sum)
names(bydate_df_sum_imputed) <- c('day', 'total_steps')
hist(bydate_df_sum_imputed$total_steps, xlab = 'Day Number', main = 'Total Steps per Day (on Imputed Data)', 
     col = 10, breaks = 50)
```

##Are there differences in activity patterns between weekdays and weekends? (Question 5)


```{r}
#creating requisite date/day category variables
mean_imp_df$date <- ymd(mean_imp_df$date)

class(mean_imp_df$date)

mean_imp_df$dayofweek <- weekdays(mean_imp_df$date)
head(mean_imp_df)

mean_imp_df$day_category <- ifelse(mean_imp_df$dayofweek %in% c("Saturday", "Sunday"),"Weekend", "Weekday")

#double checking
head(mean_imp_df)

#aggregating to the interval level
mean_imp_byinterval <- aggregate(steps ~ interval + day_category, mean_imp_df, mean)

head(mean_imp_byinterval)
```

###Generating the panel line plot 
```{r}
ggplot(mean_imp_byinterval, aes(x=interval,y=steps)) + 
  geom_line(color="red",size=1) + facet_wrap(~day_category, nrow=2, ncol=1) + labs(x="Interval",y="Number of Steps")
```

